---
title: "Federal Public Investment in Mexico, 1959-1963"
subtitle: "Descriptive statistics — institutions and states (Cuadros 15–23, *Inversión Pública Federal*)"
author: "TapadosPRI"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_caption: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: flatly
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 5.2, dpi = 110
)

library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)

# Choropleth maps need `sf`; if it isn't installed the map chunks are skipped
# gracefully (the rest of the report still renders).
has_sf <- requireNamespace("sf", quietly = TRUE)
if (has_sf) library(sf)

theme_set(theme_minimal(base_size = 13) +
            theme(panel.grid.minor = element_blank(),
                  plot.title = element_text(face = "bold")))

# Format-robust table helper (renders in both PDF/LaTeX and HTML)
ktab <- function(x, caption = NULL) {
  kableExtra::kable_styling(
    knitr::kable(x, caption = caption, booktabs = TRUE, linesep = ""),
    full_width = FALSE,
    latex_options = c("striped", "HOLD_position"),
    bootstrap_options = c("striped", "condensed"))
}

# --- Locate the Dropbox data root the same way config.py does ------------------
read_db_root <- function() {
  root <- Sys.getenv("TAPADOSPRI_DB_ROOT")
  if (nzchar(root)) return(path.expand(root))
  # fall back to parsing the repo .env (…/code/05-investment/plan/ -> repo root)
  for (up in 1:5) {
    envf <- file.path(paste(rep("..", up), collapse = "/"), ".env")
    if (file.exists(envf)) {
      ln <- grep("^TAPADOSPRI_DB_ROOT=", readLines(envf), value = TRUE)
      if (length(ln)) return(path.expand(sub("^[^=]*=", "", ln[1])))
    }
  }
  path.expand("~/Dropbox/TapadosPRI")
}
DB   <- read_db_root()
DATA <- file.path(DB, "data")
INV  <- file.path(DATA, "investment")
```

```{r load-data}
long <- read_csv(file.path(INV, "federal_investment_long.csv"), show_col_types = FALSE)

# Line items = actual institution × state values (subtotals/grand totals excluded)
items <- long %>% filter(row_type == "line_item", !is.na(investment_mdp))

sector_lab <- c(
  gobierno_federal              = "Gobierno Federal",
  organismos_descentralizados   = "Organismos Descentralizados",
  empresas_participacion_estatal = "Empresas de Participación Estatal"
)
items <- items %>% mutate(sector_lab = factor(sector_lab[sector], levels = sector_lab))

# --- Year -> president lookup (Mexican sexenios). Attribution: a fiscal year is
# assigned to whoever governed during it; the Dec-1 handovers mean 1959–1963 are
# entirely Adolfo López Mateos. The table generalizes if earlier years are added.
presidents <- tribble(
  ~start, ~end, ~president,                ~party,
  1946,   1952, "Miguel Alemán Valdés",    "PRI",
  1952,   1958, "Adolfo Ruiz Cortines",    "PRI",
  1958,   1964, "Adolfo López Mateos",     "PRI",
  1964,   1970, "Gustavo Díaz Ordaz",      "PRI"
)
year_president <- function(y) {
  presidents %>% filter(y > start, y <= end) %>% slice(1) %>%
    pull(president) %>% (\(x) if (length(x)) x else NA_character_)()
}
items <- items %>% mutate(president = map_chr(year, year_president))

# --- Deflate to REAL pesos (constant 1960 = 100) ------------------------------
# Deflator built by 02_build_deflator.py from the World Bank CPI series; 1959 is
# assigned the 1960 value. Real = nominal * 100 / deflator. Downstream code sums
# `investment_mdp`, so we overwrite it with the real value and keep the nominal.
defl <- read_csv(file.path(INV, "price_deflator_mexico.csv"), show_col_types = FALSE)
items <- items %>%
  left_join(defl %>% select(year, deflator, inflation_pct), by = "year") %>%
  mutate(investment_nominal = investment_mdp,
         investment_mdp      = investment_nominal * 100 / deflator)

# Canonical state -> GeoJSON `name` (mirrors config.py STATE_TO_GEOJSON)
state_to_geojson <- c(
  "Aguascalientes"="Aguascalientes","Baja California"="Baja California",
  "Baja California Sur"="Baja California Sur","Campeche"="Campeche",
  "Chiapas"="Chiapas","Chihuahua"="Chihuahua","Coahuila"="Coahuila",
  "Colima"="Colima","Durango"="Durango","Federal District"="Ciudad de México",
  "Guanajuato"="Guanajuato","Guerrero"="Guerrero","Hidalgo"="Hidalgo",
  "Jalisco"="Jalisco","Mexico"="México","Michoacan"="Michoacán",
  "Morelos"="Morelos","Nayarit"="Nayarit","Nuevo Leon"="Nuevo León",
  "Oaxaca"="Oaxaca","Puebla"="Puebla","Queretaro"="Querétaro",
  "Quintana Roo"="Quintana Roo","San Luis Potosi"="San Luis Potosí",
  "Sinaloa"="Sinaloa","Sonora"="Sonora","Tabasco"="Tabasco",
  "Tamaulipas"="Tamaulipas","Tlaxcala"="Tlaxcala","Veracruz"="Veracruz",
  "Yucatan"="Yucatán","Zacatecas"="Zacatecas"
)
```

# Overview {-}

This report summarises **federal public investment** recorded in the yearly cuadros
*"Inversión de las dependencias, organismos y empresas del sector público, por
entidades federativas"* for **`r min(items$year)`–`r max(items$year)`**.
Each observation is one institution's investment in one state in one year.
All figures are reported in **real *millones de pesos*, constant 1960 prices**,
deflated with the World Bank consumer price index (see below).

> **Deflation.** Nominal pesos are converted to constant 1960 pesos using the
> Mexican CPI (World Bank WDI), rebased so 1960 = 100. Because the series begins
> in 1960, fiscal year 1959 is assigned the 1960 price level. Inflation was low in
> this window (about 3.4% cumulative 1960-1963), so the real adjustment is modest but
> makes the yearly totals directly comparable and consistent with the eventual
> 1925–2000 panel.

> **Presidency note.** All five fiscal years fall within a single sexenio —
> **Adolfo López Mateos (PRI, Dec 1958 – Nov 1964)**. "By presidency" cuts below
> therefore describe López Mateos' term; the code carries a general year-to-president
> map so the same tables extend automatically if earlier cuadros (back to 1925)
> are added.

```{r overview-tab}
tibble(
  Metric = c("Years", "Institutions (distinct)", "States", "Sectors",
             "Institution × state records", "Total investment (real mdp, 1960)"),
  Value  = c(
    paste(min(items$year), "–", max(items$year)),
    comma(n_distinct(items$institution)),
    as.character(n_distinct(items$state)),
    as.character(n_distinct(items$sector)),
    comma(nrow(items)),
    comma(round(sum(items$investment_mdp)))
  )
) %>% ktab(caption = "Coverage of the extracted database")
```

# Total federal investment by year

```{r total-by-year}
by_year <- items %>% group_by(year, president) %>%
  summarise(real = sum(investment_mdp),
            nominal = sum(investment_nominal), .groups = "drop")

ggplot(by_year, aes(year, real)) +
  geom_col(aes(fill = president), width = 0.7) +
  geom_text(aes(label = comma(round(real))), vjust = -0.5, size = 3.4) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Total federal public investment by year",
       subtitle = "Real millones de pesos (constant 1960 prices). Colour = president in office.",
       x = NULL, y = "Investment (real mdp, 1960)", fill = "President")
```

```{r nominal-vs-real, fig.height=4}
by_year %>%
  pivot_longer(c(real, nominal), names_to = "measure", values_to = "total") %>%
  mutate(measure = recode(measure, real = "Real (1960 pesos)", nominal = "Nominal")) %>%
  ggplot(aes(year, total, colour = measure)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_y_continuous(labels = comma) +
  scale_colour_manual(values = c("Nominal" = "grey55", "Real (1960 pesos)" = "#d95f02")) +
  labs(title = "Nominal vs. real federal investment",
       subtitle = "The gap widens as prices drift up over the sexenio",
       x = NULL, y = "Investment (mdp)", colour = NULL)
```

```{r sector-by-year}
items %>% group_by(year, sector_lab) %>%
  summarise(total = sum(investment_mdp), .groups = "drop") %>%
  ggplot(aes(year, total, fill = sector_lab)) +
  geom_col(position = "fill", width = 0.7) +
  scale_y_continuous(labels = percent) +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "Composition of investment by sector",
       subtitle = "Share of yearly total across the three public-sector families",
       x = NULL, y = "Share of total", fill = "Sector")
```

# Rankings

## Top institutions overall

```{r top-institutions}
top_inst <- items %>% group_by(institution, sector_lab) %>%
  summarise(total = sum(investment_mdp), .groups = "drop") %>%
  slice_max(total, n = 12)

ggplot(top_inst, aes(reorder(institution, total), total, fill = sector_lab)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "Top 12 institutions by total investment, 1959-1963",
       x = NULL, y = "Investment (real mdp, 1960)", fill = "Sector")
```

## Top 10 states overall

```{r top-states}
by_state <- items %>% group_by(state) %>%
  summarise(total = sum(investment_mdp), .groups = "drop") %>%
  arrange(desc(total))

by_state %>% slice_max(total, n = 10) %>%
  ggplot(aes(reorder(state, total), total)) +
  geom_col(fill = "#2c7fb8") +
  geom_text(aes(label = comma(round(total))), hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Top 10 states by total federal investment, 1959-1963",
       subtitle = "Real millones de pesos (constant 1960 prices)",
       x = NULL, y = "Investment (real mdp, 1960)")
```

## Top 5 states and institutions by presidency

```{r top-by-presidency}
# One row per rank (PDF-friendly). All years fall in the López Mateos sexenio.
st5 <- items %>% group_by(state) %>%
  summarise(t = sum(investment_mdp), .groups = "drop") %>%
  slice_max(t, n = 5) %>%
  transmute(Rank = row_number(), State = state, `State (real mdp)` = comma(round(t)))

in5 <- items %>% group_by(institution) %>%
  summarise(t = sum(investment_mdp), .groups = "drop") %>%
  slice_max(t, n = 5) %>%
  transmute(Rank = row_number(), Institution = institution,
            `Inst. (real mdp)` = comma(round(t)))

left_join(st5, in5, by = "Rank") %>%
  ktab(caption = sprintf("Top 5 states and institutions, %s sexenio (%d–%d)",
                         unique(items$president), min(items$year), max(items$year)))
```

# State analysis {.tabset}

The volume is organised *por entidades federativas*, so the state dimension is its
strongest feature. This section maps where federal money went, how concentrated it
was, and how state allocations shifted across the sexenio.

## Choropleth — where the money went

```{r map, fig.height=6, eval=has_sf}
geo <- st_read(file.path(DATA, "shapefiles", "mexico_states.json"), quiet = TRUE)

state_tot <- by_state %>% mutate(geo_name = state_to_geojson[state])
geo2 <- geo %>% left_join(state_tot, by = c("name" = "geo_name"))

ggplot(geo2) +
  geom_sf(aes(fill = total), colour = "white", linewidth = 0.2) +
  scale_fill_viridis_c(labels = comma, option = "C", na.value = "grey90",
                       trans = "sqrt") +
  labs(title = "Total federal investment by state, 1959-1963",
       subtitle = "Millones de pesos (sqrt colour scale). Grey = unmatched.",
       fill = "real mdp") +
  theme_void(base_size = 12)
```

Because the **Federal District** and a few industrial states dominate, the map
below drops the Federal District to reveal variation across the rest of the country.

```{r map-no-df, fig.height=6, eval=has_sf}
geo3 <- geo2 %>% mutate(total = if_else(name == "Ciudad de México", NA_real_, total))
ggplot(geo3) +
  geom_sf(aes(fill = total), colour = "white", linewidth = 0.2) +
  scale_fill_viridis_c(labels = comma, option = "C", na.value = "grey85") +
  labs(title = "Total federal investment by state (excl. Federal District)",
       fill = "real mdp") +
  theme_void(base_size = 12)
```

## State × year heatmap

```{r heatmap, fig.height=7}
items %>% group_by(state, year) %>%
  summarise(total = sum(investment_mdp), .groups = "drop") %>%
  mutate(state = fct_reorder(state, total, .fun = sum)) %>%
  ggplot(aes(factor(year), state, fill = total)) +
  geom_tile(colour = "white") +
  scale_fill_viridis_c(labels = comma, option = "D", trans = "sqrt") +
  labs(title = "Federal investment by state and year",
       subtitle = "States ordered by cumulative investment (sqrt colour scale)",
       x = NULL, y = NULL, fill = "real mdp")
```

## Sector mix by state

```{r state-sector, fig.height=6}
items %>%
  semi_join(by_state %>% slice_max(total, n = 15), by = "state") %>%
  group_by(state, sector_lab) %>%
  summarise(total = sum(investment_mdp), .groups = "drop") %>%
  mutate(state = fct_reorder(state, total, .fun = sum)) %>%
  ggplot(aes(total, state, fill = sector_lab)) +
  geom_col(position = "fill") +
  scale_x_continuous(labels = percent) +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "Sector composition of investment, top 15 states",
       subtitle = "Federal ministries vs. decentralised organisms vs. state enterprises",
       x = "Share of state total", y = NULL, fill = "Sector")
```

## Growth across the sexenio

```{r growth, fig.height=6}
growth <- items %>% filter(year %in% c(min(year), max(year))) %>%
  group_by(state, year) %>% summarise(total = sum(investment_mdp), .groups = "drop") %>%
  pivot_wider(names_from = year, values_from = total, names_prefix = "y") %>%
  rename(y0 = 2, y1 = 3) %>%
  mutate(change = y1 - y0) %>% arrange(desc(change))

growth %>%
  ggplot(aes(change, reorder(state, change), fill = change > 0)) +
  geom_col() +
  scale_x_continuous(labels = comma) +
  scale_fill_manual(values = c(`TRUE` = "#1a9850", `FALSE` = "#d73027"), guide = "none") +
  labs(title = sprintf("Change in state investment, %d vs %d",
                       min(items$year), max(items$year)),
       subtitle = "Real millones de pesos (constant 1960 prices)",
       x = "Change in real investment (mdp, 1960)", y = NULL)
```

## Concentration & inequality

```{r concentration}
gini <- function(x) {
  x <- sort(x[!is.na(x)]); n <- length(x)
  if (n < 2 || sum(x) == 0) return(NA_real_)
  sum((2 * seq_len(n) - n - 1) * x) / (n * sum(x))
}
conc <- function(d) {
  d <- arrange(d, desc(total))
  tibble(
    `Top-1 share`  = percent(d$total[1] / sum(d$total), 0.1),
    `Top-5 share`  = percent(sum(d$total[1:5]) / sum(d$total), 0.1),
    `HHI`          = round(sum((d$total / sum(d$total))^2), 3),
    `Gini`         = round(gini(d$total), 3)
  )
}
bind_rows(
  conc(by_state) %>% mutate(Sample = "All states", .before = 1),
  conc(by_state %>% filter(state != "Federal District")) %>%
    mutate(Sample = "Excl. Federal District", .before = 1)
) %>% ktab(caption = "Geographic concentration of federal investment")
```

```{r lorenz, fig.height=4.6, fig.width=6}
lor <- by_state %>% arrange(total) %>%
  mutate(p_states = row_number() / n(),
         p_inv    = cumsum(total) / sum(total))
ggplot(lor, aes(p_states, p_inv)) +
  geom_abline(linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 1, colour = "#2c7fb8") +
  geom_point(size = 1.5, colour = "#2c7fb8") +
  scale_x_continuous(labels = percent) + scale_y_continuous(labels = percent) +
  labs(title = "Lorenz curve: investment across states",
       subtitle = "Distance from the 45° line = geographic inequality",
       x = "Cumulative share of states", y = "Cumulative share of investment")
```

# Infrastructure view — excluding PEMEX and the Federal District Department

Two line items dominate and distort the geography: **Petróleos Mexicanos** (oil
investment, concentrated in a handful of producing states) and the **Departamento
del Distrito Federal** (Mexico City's own municipal budget). Together they are
**about 27% of all investment**. Removing both isolates cross-state federal
*infrastructure* — roads, irrigation, electrification, schools, hospitals, ports.

```{r infra-data}
excl  <- c("Petróleos Mexicanos", "Departamento del Distrito Federal")
infra <- items %>% filter(!institution %in% excl)
infra_state <- infra %>% group_by(state) %>%
  summarise(total = sum(investment_mdp), .groups = "drop")
```

```{r infra-year, fig.height=4.4}
bind_rows(
  items %>% group_by(year) %>% summarise(total = sum(investment_mdp), .groups = "drop") %>%
    mutate(view = "All institutions"),
  infra %>% group_by(year) %>% summarise(total = sum(investment_mdp), .groups = "drop") %>%
    mutate(view = "Excl. PEMEX & DDF")
) %>%
  ggplot(aes(year, total, colour = view)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_colour_manual(values = c("All institutions" = "grey55",
                                 "Excl. PEMEX & DDF" = "#1b9e77")) +
  labs(title = "Federal investment: all institutions vs. infrastructure only",
       subtitle = "Real millones de pesos (1960). PEMEX and the DDF removed.",
       x = NULL, y = "Investment (real mdp, 1960)", colour = NULL)
```

```{r infra-top-states, fig.height=5}
infra_state %>% slice_max(total, n = 10) %>%
  ggplot(aes(reorder(state, total), total)) +
  geom_col(fill = "#1b9e77") +
  geom_text(aes(label = comma(round(total))), hjust = -0.1, size = 3.2) +
  coord_flip() +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Top 10 states by infrastructure investment (excl. PEMEX & DDF)",
       subtitle = "Real millones de pesos (constant 1960 prices)",
       x = NULL, y = "Investment (real mdp, 1960)")
```

```{r infra-map, fig.height=6, eval=has_sf}
istate <- infra_state %>% mutate(geo_name = state_to_geojson[state])
geo_i  <- geo %>% left_join(istate, by = c("name" = "geo_name"))
ggplot(geo_i) +
  geom_sf(aes(fill = total), colour = "white", linewidth = 0.2) +
  scale_fill_viridis_c(labels = comma, option = "C", na.value = "grey90", trans = "sqrt") +
  labs(title = "Infrastructure investment by state (excl. PEMEX & DDF)",
       subtitle = "Real millones de pesos (sqrt colour scale). Grey = Federal District / unmatched.",
       fill = "real mdp") +
  theme_void(base_size = 12)
```

The share each state loses when PEMEX and the DDF are removed reveals who the
"headline" numbers really flattered:

```{r infra-shift}
full_state <- by_state %>% rename(all = total)
infra_state %>% rename(infra = total) %>%
  right_join(full_state, by = "state") %>%
  mutate(infra = coalesce(infra, 0),
         removed_mdp = all - infra,
         removed_pct = removed_mdp / all) %>%
  slice_max(removed_pct, n = 8) %>%
  transmute(State = state,
            `All (real mdp)` = comma(round(all)),
            `Infra only` = comma(round(infra)),
            `Removed %` = percent(removed_pct, 0.1)) %>%
  ktab(caption = "States most reduced by removing PEMEX and the DDF")
```

# Recommended further analysis {-}

- **Deflator robustness.** Figures are now real (CPI, 1960 = 100). A wholesale or
  gross-fixed-capital-formation deflator (INEGI/Banxico) is arguably better for
  *investment* than the CPI and would be a useful sensitivity check; the WPI series
  here is empty before ~1970, so it can only anchor the later panel.
- **Per-capita and per-area allocation.** Join 1960-census state population to see
  whether high totals (Federal District, Coahuila) survive normalisation, or whether
  poorer states received more *per head* — the real distributive question.
- **Extend the panel to 1925–1963.** The full volume covers earlier cuadros; adding
  them turns the president lookup into a genuine cross-sexenio comparison
  (Cárdenas, Ávila Camacho, Alemán, Ruiz Cortines, López Mateos).
- **Volatility of allocations.** Coefficient of variation of each state's yearly
  share flags politically "swing" vs. stable recipients.
- **Link to the biography networks (project payoff).** Cross the leading
  institutions here (SOP, SRH, PEMEX, CFE) with the `govt_positions` careers: did
  tapados and their network alters cluster in the highest-investment ministries?
  Use `total_mdp` from `federal_investment_institutions.csv` as a **budget-based
  institution-importance weight** for the manual rank tables in `data/rank/`.
- **Regional politics.** Aggregate to regions (North, Bajío, Centre, South-East)
  and test whether investment tracked governors' alignment with the president.

```{r sessioninfo, echo=FALSE}
cat("Data source:", INV, "\nRendered:", format(Sys.time()))
```
